Distributional Trees and Forests for Circular Data with an Application to Probabilistic Wind Direction Forecasting

### 2019-10-18 pkg/circtree/inst/wind_paper/draft/draft.Rnw
added references and information on CRPS skill score, restructured introduction and application section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2801 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: dad6a845b2a7ac82a1cdd7d220fa4b8048b0adb3

\title{Distributional Trees and Forests for Circular Data}
%\Shorttitle{Distributional Trees and Forests for Circular Data}

@@

In many approaches to model circular data it is assumed that the circular variable of interest 
follows a circular distribution, in particular the von Mises distribution (which will be discussed 
in more detail in Section~\ref{sec:prob_circ}). This was also the basis on which early regression
models for circular data were built on as they regress the distribution parameter(s) on covariates.
One of the first regression models with a circular response variable and linear covariates
was presented by \cite{Gould:1969} where a linear combination of the covariates is employed to model
the circular mean, hence the location parameter of the von Mises distribtuion. 
\cite{Johnson+Wehrly:1978} took this idea one step further 
by plugging in a link function such that the values of the linear predictor are transformed 
to an restricted interval of length $2\pi$. 
This model was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
introducing another specific transformation function which maps the values of the real line to the 
unit circle. Moreover, the presented methodology allows for the mean direction or the dispersion 
to be modeled not only in separate models but also in a mixed model. However, applying these models 
has shown to be very challenging, mainly due to the complexity encountered in optimizing the 
log-likelihood function.
Additionally, the named approaches and also its further developments 
\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001, Mulder+Klugkist:2017}
are built on additive regression models which may not be able to capture all types of effects.
For example abrupt changes might not be detected properly owing to the fixed structure enforced
by model assumptions such as the type of link function and the type of linear combination.
Therefore, we propose an adaption of regression trees and random forests to circular data by employing 
distributional trees and distributional forests. 
In that way no model assumptions are required as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while combining an 
ensemble of trees to a forest model allows for modeling smooth changes as well. 
In that way, a high level of flexibility is provided while a full distributional model is specified 
by distributional trees and forests such that a wide range of inference is made available.
These features and the corresponding methodology are presented in Section~\ref{sec:tree_forest} after discussing 
distributional models for circular data in Section~\ref{sec:prob_circ}.
In Section~\ref{sec:wind} the presented models a applied to probabilistic forecasting of wind direction
at the airports of Innsbruck and Vienna.

@@

\begin{figure}[t]
\centering
\includegraphics[width=4cm]{density_vM.pdf}
\caption{Empirical histogram (gray) and fitted density (red line)
are depicted along with the estimated location parameter (red hand).
\fixme{refer to data, Innsbruck airport}
}

\label{fig:density_vM} 
\end{figure}

Figure~\ref{fig:density_vM} illustrates a fitted distributional model employing
the von Mises distribution for observed wind direction data from the airport of Innsbruck.
This example shows that reasonably well fitting probabilistic models can
be obtained for circular data by specifying an appropriate distribution family 
such as the von Mises distribution and estimating its parameters via maximum likelihood. 
However, this procedure considers only the response variable $Y$. Including covariates 
can improve any model remarkably as they might provide valuable additional information.
Therefore, distributional regression models are of high interest for all types of data,
including circular data.

\section{Regression Trees and Forests for Circular Data}
\label{sec:tree_forest}
\subsection{Motivation}
A commonly used approach to including covariates as regressors in distributional models is 
provided by generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
There, each distribution parameter is linked to a linear predictor by an additive function.
For circular data the mixed model by \cite{Fisher:1993} follows the idea of modeling both
parameters of the von Mises distribution, however, in an iterative optimization procedure
requiring well fitting initial values for the parameters that are recommended to be calculated 
in advance in two separate models. Using the obtained initial value the model parameters are 
estimated and improved stepwise until convergence. While providing a well elaborated theory 
this methodology and the contained optimization problem have shown to be very complex and 
bear the risk of not converging if not applied thoughtfully. \fixme{citation} %\citealp{}.
Furthermore, these additive strategies work well for smooth effects but difficulties can arise 
in case of non-additive changes. Additionally, specifying a model properly can be very challenging,
particularly for a high number of covariates and no information about possible interactions.
An alternative regression approach that can deal with these problems is to apply a tree-structured 
model.

@@

\section{Wind Direction Forecasts}
\label{sec:wind}
\subsection{Motivation}
Wind is a classical circular quantity and accurate forecasts of wind direction
are of great importance for decision-making processes and risk management,
e.g., in air traffic management or renewable energy production.

@@

As a motivational example Figure~\ref{fig:tree_ibk} illustrates a distributional
tree applied to predict the wind direction at the airport of Innsbruck in the following hour. 
Innsbruck lies at the bottom of a deep valley in the Alps. Topography channels wind along the 
west-east valley axis or along a tributary valley intersecting from the south. Hence, pressure 
gradients to which valley wind regimes react both west and east of the airport are considered 
as covariates along with other meteorological measurements at the airport (lagged by one hour),
such as wind direction and wind speed at Innsbruck Airport.

@@

\subsection{Evaluation Setting and Models Applied}
Along with distributional trees and forests for circular data three other models have been applied
in this study. In particular, they have been evaluated in a five-fold cross validation with each of 
the five years contained in the data set being used as testing data once after the models where 
learned on the remaining four years. 
In order to compare the performances of the employed models a circular version of the continuous 
ranked probability score (CRPS) as introduced by \cite{Grimit+Gneiting+Berrocal:2006} is calculated.
As the linear version of the CRPS it measures the difference between an observation
and the corresponding predicted distribution function in order to assess the probabilistic goodness 
of fit. In order to adjust this general methodology to circular data an additional auxiliary function
is employed which measures the angular distance between two circular variables. 
The resulting values are summed up over each observation contained in the learning data and the 
corresponding predicted distribution parameters of each of the models.

@@

\paragraph{Generalized Linear Model}
Another commonly applied approach is to regress the parameter(s) of the von Mises distribution
on covariates by a linear combination of selected covariates connected to the parameter(s) via
a specified link function. The toolbox of generalized linear models ...

\fixme{information on applied implementation}

@@

\section{Discussion and outlook}

@@

\subsection{Splits in circular covariates}

In order to obtain more parsimonious and more stable trees another possible
extension for \emph{circular covariates} (with or without a \emph{circular response})
is to consider their circular nature when searching the best split into two segments.
In general, searching the best separation of a covariate into a ``left'' and ``right''
daughter node tries to maximize the segmented log-likelihood:
\begin{equation}
\max \left(\sum_{y \in \mathit{left}} \ell(\hat{\mu_1}, \hat{\kappa_1}; y) + \sum_{y \in \mathit{right}} \ell(\hat{\mu_2}, \hat{\kappa_2}; y)\right)
\end{equation}
where $\hat{\mu_1}$, $\hat{\kappa_1}$, $\hat{\mu_2}$, $\hat{\kappa_2}$ are the estimated parameters
of the von Mises distribution in the two daughter nodes. Searching a single split point $\nu$ in a circular covariate $\in [0, 2 \pi)$
only considers linear splits into the intervals $\mathit{left}=[0,\nu]$ and $\mathit{right}=(\nu,2\pi)$,
thus enforcing a potentially unnatural separation at zero. This can be avoided by searching
for two split points $\nu$ and $\tau$ considering a split into one interval $\mathit{left}=[\nu,\tau]$
and its complement $\mathit{right}=[0,\nu) \cup (\tau,2\pi)$, encompassing zero. The latter
strategy is invariant to the (often arbitrary) definition of the direction at zero.

When one split point $\nu$ is sufficiently close to zero and the other $\tau$ sufficiently
far away, a simple linear split typically suffices to capture such a split (as seen for the
lagged wind direction in Figure~\ref{fig:tree_ibk}). If both $\nu$ and $\tau$ differ
clearly from zero, two linear splits should also lead to a reasonable (but less parsimonious)
fit. However, if both $\nu$ and $\tau$ are rather close to zero, a linear split strategy
might miss such a pattern.

The required test statistic to maximally select two split points simultaneously is straightforward
to accommodate in the CTree framework by providing all binary indicators corresponding
to the splits into $\mathit{left}$/$\mathit{right}$ intervals. However, this will become
increasingly slow for larger sample sizes but it might be possible to speed up computations by
exploiting the particular covariance structure similar to Hothorn and Zeileis~(2008). In the
MOB framework the extension is not quite as straightforward but one strategy could be to
adapt double maximum tests \`a la \cite{Bai+Perron:2003}.

Hence, the splitting idea can be naturally extended to a two-point search, however, for 
an unbiased and inference-based selection the corresponding testing strategies might need
further adaption.

>>>>

\title{Distributional Trees and Forests for Circular Data with an Application to Probabilistic Wind Direction Forecasting}
\Shorttitle{Distributional Trees and Forests for Circular Data}

@@

In many approaches to model circular data it is assumed that the circular variable of interest 
follows a circular distribution, in particular the von Mises distribution which is also known 
as the ``circular normal distribution'' (see Section~\ref{sec:prob_circ} for more details). 
This was also the basis which early regression models for circular data were built on as they 
regress the distribution parameter(s) on covariates.

One of the first regression models with a circular response variable and linear covariates
was presented by \cite{Gould:1969} where the circular mean, hence the location parameter of 
the von Mises distribtuion, is predicted by a linear combination of covariates.
\cite{Johnson+Wehrly:1978} took this idea one step further by employing generalized linear 
model (GLM)-type approach, i.e., by plugging in a link function such that the values of the 
linear predictor are transformed to a restricted interval of length $2\pi$. 
This model was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
introducing another specific transformation function which maps the values of the real line to the 
unit circle. Moreover, the presented methodology allows for both distribution parameters, the mean direction and the concentration parameter, to be modeled not only in separate models but also in a 
mixed model. There, within an iterative optimization algorithm the two parameters are estimated and 
improved stepwise until convergence. To obtain proper initial values it is recommended to apply
two separate models where either one of the parameters is fixed while the other is modeled depending
on covariates. While providing a well elaborated theory this methodology and the contained optimization 
problem have shown to be very challenging, mainly due to the complexity encountered in optimizing the 
log-likelihood function that is not globally concave, and thus bear a high risk of not converging if 
not applied thoughtfully \citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}.
An alternative approach to regression modeling of circular data that avoids this strong dependence 
on choosing well-fitting initial values is presented by \cite{Mulder+Klugkist:2017} employing a GLM 
in a Bayesian framework.

The named methods and also many of their further developments 
\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive regression models.
While they provide a powerful toolbox, particularly for modeling smooth effects, they may not be 
able to capture all types of effects. For example, abrupt changes might not be detected properly owing 
to the fixed structure enforced by model assumptions such as the type of link function and the 
form of linear combinations.

Therefore, we propose a very flexible alternative approach by adapting regression trees and random 
forests to circular data, specifically by employing distributional trees and distributional forests
following \cite{Schlosser+Hothorn+Stauffer:2019}. 
In that way no model assumptions are required as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while combining an 
ensemble of trees to a forest model allows for modeling smooth changes as well. 
In that way, a high level of flexibility is provided while a full distributional model is specified 
by distributional trees and forests such that a wide range of inference is made available.
These features and the corresponding methodology are presented in Section~\ref{sec:tree_forest} after
discussing distributional models for circular data in Section~\ref{sec:prob_circ}.
In Section~\ref{sec:wind} the presented models a applied to probabilistic forecasting of wind direction
at the airports of Innsbruck and Vienna.

@@

\begin{figure}[t]
\centering
\includegraphics[width=4cm]{density_vM.pdf}
\caption{Empirical histogram (gray) and fitted density (red line)
are depicted along with the estimated location parameter (red hand).
}
\label{fig:density_vM} 
\end{figure}

Figure~\ref{fig:density_vM} illustrates a distributional model fitted to circular data
employing the von Mises distribution.
% on observed wind direction data from the airport of Innsbruck.
This example shows that reasonably well fitting probabilistic models can
be obtained for circular data by specifying an appropriate distribution family 
such as the von Mises distribution and estimating its parameters via maximum likelihood. 
However, this procedure considers only the response variable $Y$. Including covariates 
can improve any model remarkably as they might provide valuable additional information.
Therefore, distributional regression models are of high interest for all types of data,
including circular data.

\section{Regression Trees and Forests for Circular Data}
\label{sec:tree_forest}
%\subsection{Motivation}
A common way to including covariates as regressors in distributional models is to apply
generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} where each distribution parameter is estimated
by its own generalized additive model (GAM).
As already discussed in Section~\ref{sec:introduction}, additive strategies work well 
for smooth effects but difficulties can arise in case of non-additive changes. Additionally, 
specifying a model properly can be very challenging, particularly for a high number of covariates 
and no information about possible interactions.
An alternative regression approach that can deal with these problems is to apply a tree-structured 
model.

@@

\section{Case Study: Wind Direction Forecasts}
\label{sec:wind}

Wind is a classical circular quantity and accurate forecasts of wind direction
are of great importance for decision-making processes and risk management,
e.g., in air traffic management or renewable energy production.

@@

\subsection{Motivational Example}
As a first illustration Figure~\ref{fig:tree_ibk} displays a distributional
tree applied to predict the wind direction at the airport of Innsbruck in the following hour. 
Innsbruck lies at the bottom of a deep valley in the Alps. Topography channels wind along the 
west-east valley axis or along a tributary valley intersecting from the south. Hence, pressure 
gradients to which valley wind regimes react both west and east of the airport are considered 
as covariates along with other meteorological measurements at the airport (lagged by one hour),
such as wind direction and wind speed at Innsbruck Airport.

@@

\subsection{Evaluation Setting and Models Applied}
Along with distributional trees and forests for circular data three other models have been applied
in this study. In particular, they have been evaluated in a five-fold cross validation with each of 
the five years contained in the data set being used as testing data once after the models where 
learned on the remaining four years. 
In order to compare the performances of the employed models a circular version of the continuous 
ranked probability score (CRPS) as introduced by \cite{Grimit+Gneiting+Berrocal:2006} is calculated.
Just as the linear version of the CRPS \citep[see][for more details]{Hersbach:2000} it measures the 
difference between an observation and the corresponding predicted distribution function in order 
to assess the probabilistic goodness of fit of the assessed model. Hence, the lower the CRPS value 
the better the performance. In order to adjust the general methodology of the CRPS to circular data 
\cite{Grimit+Gneiting+Berrocal:2006} employ an auxiliary function which measures the angular distance 
between two circular variables. 
The circular CRPS is evaluated out of sample and the resulting values are averaged over 
each observation contained in the testing data and the corresponding predicted distribution 
parameters of each of the models.
To measure the improvement of the trees, forests, persistence and linear models 
over the climatology model the skill score based on the circular CRPS is evaluated
with the climatology model used as reference:
\begin{equation}
\text{CRPSS}_{\text{method}} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{climatology}}}.
\end{equation}

@@

\paragraph{Generalized Linear Model}
Another commonly applied approach is to linearly regress the circular variable of interest on
its lagged values. For the here discussed application this corresponds to establishing a linear model
with wind direction as the outcome variable and wind direction of the previous time point as covariate.
Following \cite{Mulder+Klugkist:2017}, the employed model plugs in a link function in order
to guarantee values on the interval $[-\pi,\pi]$ and is of the form
\begin{align}
dd = \beta_0 + g(\beta_1 \cdot dd_{lagged})
\end{align}
with $\beta_0$ being a circular intercept, $\beta_1$ the regression coefficient and 
$g$ the link function $g(x) = r \cdot \arctan(x)$ with an optional factor $r$. 
Using the corresponding implementation provided in the \textsf{R} package \textbf{circglmbayes}
by \cite{Mulder+Klugkist:2017} the function \textbf{circGLM} estimates the
intercept and regression coefficient along with the the concentration parameter $\kappa$.

\fixme{add references and further information on the estimation process within circGLM}

\fixme{information on applied implementation here or in separate chapter?}

@@

\section{Discussion and Outlook}

@@

\fixme{Summary of results}

\fixme{Possible extensions of models, use of NWP data}

%\subsection{Splits in circular covariates}

%In order to obtain more parsimonious and more stable trees another possible
%extension for \emph{circular covariates} (with or without a \emph{circular response})
%is to consider their circular nature when searching the best split into two segments.
%In general, searching the best separation of a covariate into a ``left'' and ``right''
%daughter node tries to maximize the segmented log-likelihood:
%\begin{equation}
%\max \left(\sum_{y \in \mathit{left}} \ell(\hat{\mu_1}, \hat{\kappa_1}; y) + \sum_{y \in \mathit{right}} %\ell(\hat{\mu_2}, \hat{\kappa_2}; y)\right)
%\end{equation}
%where $\hat{\mu_1}$, $\hat{\kappa_1}$, $\hat{\mu_2}$, $\hat{\kappa_2}$ are the estimated parameters
%of the von Mises distribution in the two daughter nodes. Searching a single split point $\nu$ in a %circular covariate $\in [0, 2 \pi)$
%only considers linear splits into the intervals $\mathit{left}=[0,\nu]$ and $\mathit{right}=(\nu,2\pi)$,
%thus enforcing a potentially unnatural separation at zero. This can be avoided by searching
%for two split points $\nu$ and $\tau$ considering a split into one interval $\mathit{left}=[\nu,\tau]$
%and its complement $\mathit{right}=[0,\nu) \cup (\tau,2\pi)$, encompassing zero. The latter
%strategy is invariant to the (often arbitrary) definition of the direction at zero.

%When one split point $\nu$ is sufficiently close to zero and the other $\tau$ sufficiently
%far away, a simple linear split typically suffices to capture such a split (as seen for the
%lagged wind direction in Figure~\ref{fig:tree_ibk}). If both $\nu$ and $\tau$ differ
%clearly from zero, two linear splits should also lead to a reasonable (but less parsimonious)
%fit. However, if both $\nu$ and $\tau$ are rather close to zero, a linear split strategy
%might miss such a pattern.

%The required test statistic to maximally select two split points simultaneously is straightforward
%to accommodate in the CTree framework by providing all binary indicators corresponding
%to the splits into $\mathit{left}$/$\mathit{right}$ intervals. However, this will become
%increasingly slow for larger sample sizes but it might be possible to speed up computations by
%exploiting the particular covariance structure similar to Hothorn and Zeileis~(2008). In the
%MOB framework the extension is not quite as straightforward but one strategy could be to
%adapt double maximum tests \`a la \cite{Bai+Perron:2003}.

%Hence, the splitting idea can be naturally extended to a two-point search, however, for 
%an unbiased and inference-based selection the corresponding testing strategies might need
%further adaption.
